Optimal synaptic strategies for different timescales of memory

### 2018-03-19 Paper/synapse_main.tex
separated main, sup, etc

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}
%\usepackage{graphics} for finding documentation only
\pdfsuppresswarningpagegroup=1
\usepackage{epstopdf}
\epstopdfsetup{update,suffix=-generated}
%
\IfFileExists{"C:/Progra\string~1/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Progra\string~1/Inkscape/}}{}
\IfFileExists{"/Applications/Inkscape.app/Contents/Resources/bin/inkscape"}{\newcommand{\inkscapedir}{/Applications/Inkscape.app/Contents/Resources/bin/}}{}
\providecommand{\inkscapedir}{}
\epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{%
  "\inkscapedir inkscape" -f #1 -C -A \OutputFile}
%
\usepackage{adjustbox}
\newcommand{\aligntop}[1]{\adjustbox{valign=t}{#1}}
%\usepackage[inline]{enumitem}
\newlist{myenuma}{enumerate*}{4}
\setlist[myenuma]{label=(\alph*),ref=\alph*}
\newcommand{\panel}[1]{\item\aligntop{#1}}
\graphicspath{{"Figs/"}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb,amsfonts}
\usepackage{amsthm}
%\usepackage[notref,notcite]{showkeys}
%\usepackage[centertags]{amsmath}
%\usepackage{bibunits}
%\usepackage{hyperref}
\hypersetup{bookmarksopenlevel=1,bookmarksnumbered}
\usepackage[capitalise,nameinlink]{cleveref}
%\Crefname{appendix}{App.}{Apps.}
\creflabelformat{equation}{[#2#1#3]}
\crefrangelabelformat{equation}{[#3#1#4--#5#2#6]}
%\crefformat{equation}{#2Eq.~[#1]#3}
%\crefrangeformat{equation}{Eqs.~[#3#1#4--#5#2#6]}
%\crefmultiformat{equation}{Eqs.~[#2#1#3}{,2#1#3]}{,#2#1#3}{,#2#1#3]}
%\crefrangemultiformat{equation}{Eqs.~[#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}{,#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}
\newcommand{\main}{the main paper}
\newcommand{\supp}{supporting information}
% ----------------------------------------------------------------
\newtheoremstyle{sldefinition}%
  {3pt}%space above
  {3pt}%space below
  {}%body font
  {}%indent amount
  {\bfseries}%theorem head font
  {}%theorem head punctuation
  {\newline}%space after head
  {\thmname{#1}\thmnumber{ #2}:{\bfseries\thmnote{ #3}}}%head spec
\newtheoremstyle{slplain}%
  {3pt}%space above
  {3pt}%space below
  {}%body font
  {}%indent amount
  {\bfseries}%theorem head font
  {}%theorem head punctuation
  {\newline}%space after head
  {\thmname{#1}\thmnumber{ #2}{\bfseries\thmnote{ (#3)}}: }%head spec
%
\theoremstyle{slplain}
\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
%
\theoremstyle{sldefinition}
\newtheorem{defn}{Definition}
%
\theoremstyle{remark}
\newtheorem*{notn}{Notation}
\newtheorem*{rem}{Remark}
% ----------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO DEFINITIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
\usepackage{delim}
% brackets etc.
\delimdef\prn#1{\dleft( #1 \dright)}
\delimdef\brc#1{\dleft\{ #1 \dright\}}
\delimdef\brk#1{\dleft[ #1 \dright]}
\delimdef\abs#1{\dleft\lvert #1 \dright\rvert}
\delimdef\nrm#1{\dleft\lVert #1 \dright\rVert}
\delimdef\av#1{\dleft\langle #1 \dright\rangle}
\delimdef\avc#1#2{\dleft\langle #1 \dmiddle\vert #2 \dright\rangle}
%\newenvironment{cases}{\dleft\{\begin{aligned}}{\end{aligned}\dright.}
%
% Sets
\delimdef\set#1#2{\dleft\{ #1 \dmiddle\vert #2 \dright\}}
\delimdef\cond#1#2{\!\dleft( #1 \dmiddle\vert #2 \dright)}
\delimdef\condb#1#2{\!\dleft[ #1 \dmiddle\vert #2 \dright]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][]{\left(\frac{\partial #2}{\partial {#3}}\right)_{\!\!#1}}
\newcommand{\intd}[2][]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\CL}{\mathcal{L}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\boldsymbol{\pi}}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\arow}{\boldsymbol{\xi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Optimal synaptic strategies for different timescales of memory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Subhaneil Lahiri}
\author[a]{Surya Ganguli}

\affil[a]{Department of Applied Physics, Stanford University, Stanford CA}

% Please give the surname of the lead author for the running footer
\leadauthor{Lahiri}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: sulahiri@stanford.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{synaptic plasticity | learning | memory | neuroscience}

%\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation;LTD, long term depression}

%\begin{article}
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

  Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\setboolean{displaywatermark}{false}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{article}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

Here we initiate such a general theory by proving upper bounds on the memory curve associated with any synaptic dynamical system, within the well established ideal
observer framework of \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}.
Along the way we develop principles based on first passage time theory to order the structure of synaptic dynamical systems and relate this structure to memory performance.
We summarize our main results in the \hyperref[sec:disc]{discussion section}.

@@

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}(2 f\pot f\dep) \eq \brk{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in the \supp.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \cref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \cref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \cref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[sec:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

To maximize this flux, potentiation from a weak state must be guaranteed to end in a strong state, and depression must do the reverse.
An example of such a model is shown in \cref{fig:max_init}(\ref{fig:max_init_pot},\ref{fig:max_init_dep}).
These models have a property known as ``lumpability'' (see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time).
They are completely equivalent to a two state model with transition probabilities equal to 1, as shown in \cref{fig:max_init}(\ref{fig:binary_det}).
In particular, they have the same memory curve.

@@

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\smash{\eta^+_i}$ or increasing $\smash{\eta^-_i}$.
Because $\smash{\eta^+_i+\eta^-_i=\eta}$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter.
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.%
\footnote{Note that we do not need to worry about the order of the $\eta^\pm_i$ changing during the optimization: necessary conditions for a maximum only require that there is no infinitesimal perturbation that increases the area. Therefore we need only consider an infinitesimal neighborhood of the model, in which the order will not change.\label{ft:etaorder}}

@@

First passage time theory introduced in the \hyperref[sec:order]{previous section} becomes useful because the area has a simple expression in terms of quantities introduced in \cref{eq:kemenypm} (see \supp):
%
\begin{equation}\label{eq:areaEta}
\begin{aligned}
  A &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^+_i-\eta^+_j)\\
    &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^-_j-\eta^-_i).
\end{aligned}
\end{equation}
%

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \cref{fig:perts}, see \supp\ for details.
For example, in \cref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \cref{fig:env}(\ref{fig:sticky}),%\cref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some class of perturbations (\eg in \cref{fig:perts}) to further increase the area.

@@

Now we will look at the implications of the upper bounds found in the \hyperref[sec:bounds]{previous section} for the SNR at finite times.
As argued in \cref{eq:SNReigen}, the memory curve can be written in the form
%
\begin{equation}\label{eq:SNRmodes}
  \snr(t) = \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a}.
\end{equation}
%
The upper bounds on the initial SNR, \cref{eq:max_init}, and the area, \cref{eq:max_area}, imply the following constraints on the parameters $\brc{\CI_a,\tau_a}$:
%
\begin{equation}\label{eq:modeconstr}
  \sum_a \CI_a \leq 1,
  \qquad\qquad
  \sum_a \CI_a \tau_a \leq M-1.
\end{equation}
%
We are not claiming that these are a complete set of constraints: not every set $\brc{\CI_a,\tau_a}$ that satisfies these inequalities will actually be achievable by a synaptic model.
However, any set that violates either inequality will definitely not be achievable.

@@

% \pnasbreak splits and balances the columns before the references.
% If you see unexpected formatting errors, try commenting out this line
% as it can run into problems with floats and footnotes on the final page.
%\pnasbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\bibliographystyle{pnas-new}
\bibliography{maths,neuro,markov}

%\end{article}
%\clearpage

\begin{strip}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\appendix
\pdfbookmark[0]{Supporting information}{appendix}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\begin{center}
  \textbf{\Large Supporting information}
\end{center}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here we provide more details underlying the derivations of results in \main.

\section{Continuous time Markov processes}\label{sec:ContMarkov}

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time to define notation.
It is a generalization of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

\subsection{Notation}\label{sec:not}

For any matrix $\mathbf{A}$, we define matrices $\mathbf{A}\dg$ and $\overline{\mathbf{A}}$ as
%
\begin{equation}\label{eq:dgdef}
  \mathbf{A}\dg_{ij} \equiv \delta_{ij}\mathbf{A}_{ij},
  \qquad
  \overline{\mathbf{A}} \equiv \mathbf{A}-\mathbf{A}\dg.
\end{equation}
%
We let $\onev$ denote a column-vector of ones and $\onem=\onev\onev\trans$ denote a matrix of ones.

A continuous time Markov process is described by a matrix of transitions rates, $\MM_{ij}$, from state $i$ to $j$ with row sums equal to zero ($\MM\onev=0$).
The probabilities of being in each state at time $t$, the row-vector $\mathbf{p}(t)$, evolve according to
%
\begin{equation}\label{eq:rowsum}
  \diff{\mathbf{p}(t)}{t} = \mathbf{p}(t) \MM,
\end{equation}
%
where $\mathbf{p}(t)\onev=1$.

The equilibrium probabilities, $\eq$, satisfy
%
\begin{equation}\label{eq:equilibrium}
  \eq\MM=0,
  \qquad
  \eq\onev=1.
\end{equation}
%
As we assume an ergodic process, this eigenvalue is non-degenerate.
If all other eigenvalues have strictly negative real parts, the process is regular (aperiodic).

We define additional matrices
%
\begin{equation}\label{eq:defDLP}
  \Lb \equiv (-\MM\dg)\inv,
  \qquad
  \Pb \equiv \I + \Lb\MM.
\end{equation}
%
It can be shown that $\Lb_{ii}$ is the mean time it takes to leave state $i$ and $\Pb_{ij}$ is the probability the the next transition from state $i$ goes to state $j$:
%
\begin{equation}\label{eq:LamdaPcmpt}
  \Lb_{ii} = \frac{1}{\sum_{j \neq j} \MM_{ij}},
  \qquad
  \Pb_{ij} = %\left\{
%  \begin{aligned}
%     &0                                         && \text{if }i=j, \\
%     &\frac{\MM_{ij}}{\sum_{k \neq j} \MM_{ik}} &&\text{otherwise}.
%  \end{aligned}
%  \right.
  \begin{cases}
     0                                         &\text{if }i=j, \\
     \frac{\MM_{ij}}{\sum_{k \neq j} \MM_{ik}} &\text{otherwise}.
  \end{cases}
\end{equation}
%
Furthermore, we also define
%
\begin{equation}\label{eq:pdotD}
  \D \equiv \operatorname{diag}(\eq)\inv,
  \qquad\implies\qquad
  \eq\D=\onev\trans.
\end{equation}
%

\subsection{Fundamental matrix}\label{sec:fund}

>>>>

@@

Here we initiate such a general theory by proving upper bounds on the memory curve associated with any synaptic dynamical system, within the well established ideal
observer framework of \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}.
Along the way we develop principles based on first passage time theory to order the structure of synaptic dynamical systems and relate this structure to memory performance.
We summarize our main results in the \hyperref[sec:disc]{discussion \lcnamecref{sec:disc}}.

@@

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}(2 f\pot f\dep) \eq \brk{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in \cref{sec:SNRcurve} of the \supp.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \cref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \cref{eq:constr}?
We will derive some of these constraints in the next \lcnamecref{sec:bounds}.

@@

In the previous \lcnamecref{sec:setup}, in \cref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[eq:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

To maximize this flux, potentiation from a weak state must be guaranteed to end in a strong state, and depression must do the reverse.
An example of such a model is shown in \cref{fig:max_init}(\ref{fig:max_init_pot},\ref{fig:max_init_dep}).
These models have a property known as ``lumpability'' (see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time).
In terms of their memory curves, they are completely equivalent to a two state model with transition probabilities equal to 1, as shown in \cref{fig:max_init}(\ref{fig:binary_det}).

@@

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\smash{\eta^+_i}$ or increasing $\smash{\eta^-_i}$.
Because $\smash{\eta^+_i+\eta^-_i=\eta}$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter.
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.%
\footnote{Note that we do not need to worry about the order of the $\eta^\pm_i$ changing during the optimization: necessary conditions for a maximum only require that there is no infinitesimal perturbation that increases the area. Therefore we need only consider an infinitesimal neighborhood of the model, in which the order will not change.}%\label{ft:etaorder}}

@@

First passage time theory introduced in the \hyperref[sec:order]{previous \lcnamecref{sec:order}} becomes useful because the area has a simple expression in terms of quantities introduced in \cref{eq:kemenypm} (see \supp, \cref{sec:areadef}):
%
\begin{equation}\label{eq:areaEta}
\begin{aligned}
  A &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^+_i-\eta^+_j)\\
    &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \!\brk{\M\pot_{ij}-\M\dep_{ij}}\! (\eta^-_j-\eta^-_i).
\end{aligned}
\end{equation}
%

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \cref{fig:perts}, see \cref{sec:kuhntucker} of the \supp\ for details.
For example, in \cref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \cref{fig:env}(\ref{fig:sticky}),%\cref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some class of perturbations (\eg in \cref{fig:perts}) to further increase the area.

@@

Now we will look at the implications of the upper bounds found in the \hyperref[sec:bounds]{previous \lcnamecref{sec:bounds}} for the SNR at finite times.
As argued in \cref{eq:SNReigen}, the memory curve can be written in the form
%
\begin{equation}\label{eq:SNRmodes}
  \snr(t) = \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a}.
\end{equation}
%
The upper bounds on the initial SNR, \cref{eq:max_init}, and the area, \cref{eq:max_area}, imply the following constraints on the parameters $\brc{\CI_a,\tau_a}$:
%
\begin{equation}\label{eq:modeconstr}
  \sum_a \CI_a \leq 1,
  \qquad\qquad
  \sum_a \CI_a \tau_a \leq M-1.
\end{equation}
%
We are not claiming that these are a complete set of constraints: not every set $\brc{\CI_a,\tau_a}$ that satisfies these inequalities will actually be achievable by a synaptic model.
However, any set that violates either inequality will definitely not be achievable.

@@
